---
title: "Multivariable Regression Lab 2"
author: "Matthew Reyes"
date: '`r Sys.Date()`'
output: html_document
---

# Multivariable Regression Lab 2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 2 Notes

### Objects in R

```{r ojbects}

my_obj1 <- 48
my_obj2 <- "R is tight"
my_obj2 <- r is cool # this line won't work because of the syntax in assignment being wrong

# we can recall the object later and access the value of the object

my_obj3 <- my_obj1 + 2 # expect my_obj3 to be 50
(my_obj3)

```

### Operators and Functions

`+` addition
`-` subtraction
`/` division
`^` or `**` exponentiation
`(x mod y)` modulus, x%%y
`x %/% y` integer division

*Comparison operators*

`<` less than
`<=` less than or equal to
`>` greater than
`>=` greater than or equal to
`==` equal to
`!=` not equal to

`!` NOT
`|` OR
`&` AND

#### Common math functions

`round(pi, digits = 3)` rounds pi to 3 decimal places
`ceiling(pi)` returns the smallest integer no less than the argument
`ceiling(-pi)` returns the smallest integer no less than the argument (expect -3)
`floor(pi)` returns the largest integer no greater than the argument
`trunc(pi)` takes a single numeric argument x and returns a numeric vector containing the integers formed by truncating the values in x toward 0. `trunc(pi)` returns `3`

### Data Types and Vectors in R

*Data Types*

`class(100)` returns the class of the argument, this would return `numeric`
`is.character("100")` evalutes if the data type of the object is a character, would return true
`as.numeric("100")` would convert the character argument and return a numeric data type object, `100`
`as.character(100)` would convert the numeric argument and return a character data type object, `"100"`

*Vectors*

`c(1, 2)`, short for concatenate, constructs a vector of objects of the same data type. This would be a numeric vector
`c(1, 2, "abc")` would force the numeric arguments `1` and `2` to the character data type

Concatenate two vectors, if the vectors are different types, would convert to be the same data type

```{r concatenate_vectors}
vec1 <- c(1, 2)
vec2 <- c(4, 3)
combined <- c(vec1, vec2)
```

### Sequences

`1:5` would create a vector from 1 to 5
`5:1` would create a vector from 5 to 1
`seq(from = 1, to = 5, by = 1)` would create a vector from 1 to 5, stepping by 1
`rep(1, times = 5)` creates a vector repeting the first argument *n* times

```{r sequences}

rep(c(1,2,3), times = 3) # would create a new vector c(1,2,3, 1,2,3, 1,2,3)
rep(c(1,2,3), each = 2) # would create a new vector, repeating each element of the first argument n times. Returns c(1,1,2,2,3,3)

```

We can apply common operators and arithmetic to vectors

`c(1,2,4) * 5` returns a vector with values `c(5, 10, 20)`
`c(1,2,4) + c(2,4,3)` returns vector `c(3,6,7)`
`c(1,2,4) > 2` would check each item in the vector to see if it's greater than 2, then return `TRUE` or `FALSE` in a new vector for the evaluated condition

### Common stats functions

Given some vector x

`x <- 1:10`

`sum(x)` sum
`mean(x)` mean
`var(x)` variance
`cov(x,y)` covariance
`cor(x,y)` correlation
`sd(x)` standard dev
`length(x)` length of an item
`min(x)` min
`max(x)` max
`range(x)` range
`median(x)` sample median
`quantile(x, 0.1)` sample quantile

`na.rm = TRUE` can be used to remove NA
`summary(x)` would give a summary stats table for the vector/data frame x

`scale()` function can be used to center a vector (so that mean = 0) or standardize a vector (so that mean = 0, var = 1)

```{r scale}

x<- 1:10
y <- scale(x, center = TRUE, scale = FALSE) # center x and store output in y, y now has additional structure called "attribute"
attr(y, "scaled:center") <- NULL # removes the attribute of y

y <- c(y) # converts y to a numeric vector
z <- scale(x, center = TRUE, scale = TRUE)

```

### Indexing vectors

`c(10:1)[5]` would return the 5th element in the vector 10:1, would would be 6
`c(10:1)[c(6,3,7)]` would return the 6th, 3rd, and 7th elements in the vector 10:1 as a numeric vector containing those values
`c(10:1)[c(T,T,T,F,F,F,F,T,T,T)]` returns the elements in 10:1 for which the second vector has a TRUE value in that position
`c(10:1)[c(10:1) > 8]`, first evaluates condition `c(10:1) > 8`, providing a logical vector (TRUE, FALSE values in positions), then compares and returns a new vector in the first 10:1 vector.

`which(c(T,T,T,F,F,T))` returns the positional index of the trues in a numeric vector

`which.max()` and `which.min()` may be helpful

`sort()` sorts the values from the lowest to highest value (or sort the values form the highest to lowest if including argument `decreasing = TRUE`)
`order()` returns a permutation which rearranges the vector into ascending order (or descending)

### Matrices and Arrays

A matrix is a collection of elements stored in two dimensions (the rows and the columns). It contains elements with the same data type.

```{r matrices}

my_mat1 <- matrix(1:16, nrow = 4, ncol = 4, byrow=TRUE) # would fill in the matrix left to right
my_mat2 <- matrix(1:16, nrow = 4, ncol = 4, byrow=FALSE) # would fill in the matrix top to bottom

my_mat3 <- diag(c(1,2,3,4)) # using diag, constructs a diagonal vector, filling in 0 for the missing values

my_mat4 <- diag(2) # constructs a 2x2 identity matrix

```

```{r matrix_math}

x <- matrix(c(2,1,3,4), nrow = 2, ncol = 2, byrow = TRUE)
y <- t(x) # transposes the matrix x

X %*% y # matrix multiplication of x and y
z <- solve(x) # inverse of vector x, or x^-1

x %*% solve(x) # matrix multiplication of x and x^-1

```

Matrix element extraction

```{r extraction}

my_mat1 <- matrix(1:16, nrow=4, ncol=4, byrow=TRUE)

my_mat1[2,3] # returns the (2,3)th element (the 2nd row, 3rd col)
my_mat1[2:4, 3] # returns a vector containing the (2,3)th, (3,3)th, and (4,3)th elements of the matrix
my_mat1[2:4, 2:3] # returns a submatrix

```

### Data Frames

Arrays are a collection of elements stored in multiple dimensions. It contains elements with the same data type.

