---
title: "Multivariable Regression Lab 2"
author: "Matthew Reyes"
date: '`r Sys.Date()`'
output: html_document
---

# Multivariable Regression Lab 2

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 2 Notes

### Objects in R

```{r ojbects}

my_obj1 <- 48
my_obj2 <- "R is tight"
my_obj2 <- r is cool # this line won't work because of the syntax in assignment being wrong

# we can recall the object later and access the value of the object

my_obj3 <- my_obj1 + 2 # expect my_obj3 to be 50
(my_obj3)

```

### Operators and Functions

`+` addition
`-` subtraction
`/` division
`^` or `**` exponentiation
`(x mod y)` modulus, x%%y
`x %/% y` integer division

*Comparison operators*

`<` less than
`<=` less than or equal to
`>` greater than
`>=` greater than or equal to
`==` equal to
`!=` not equal to

`!` NOT
`|` OR
`&` AND

#### Common math functions

`round(pi, digits = 3)` rounds pi to 3 decimal places
`ceiling(pi)` returns the smallest integer no less than the argument
`ceiling(-pi)` returns the smallest integer no less than the argument (expect -3)
`floor(pi)` returns the largest integer no greater than the argument
`trunc(pi)` takes a single numeric argument x and returns a numeric vector containing the integers formed by truncating the values in x toward 0. `trunc(pi)` returns `3`

### Vectors in R















