---
title: "Multivariate Regression Lab 3"
author: "Matthew Reyes"
date: "`r Sys.Date()`"
output: html_document
---

# Multivariable Regression Lab 3

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 3 Notes

Matrices
Arrays
Graphics/plotting

### Matrices and Arrays

**Matrix** is a collection of elements stored in two dimensions. It contains elements of the *same data type*.

```{r matrices}

# `matrix()` constructs a matrix
# `byrow` indicates whether to fill by row (left to right, default = FALSE), or by column (top to bottom)

# organized by row (left to right)
my_mat1 <- matrix(1:16, nrow = 4, ncol = 4, byrow = TRUE) 

# organized by column
my_mat2 <- matrix(1:16, nrow = 4, ncol = 4, byrow = FALSE)

# `diag( c(1,2,3,4) )` would create a diagonal matrix
my_mat3 <- diag(c(1,2,3,4))

# or, create a 2x2 identity matrix using `diag(2)`
my_mat4 <- diag(2)

```

- common operators and functions can be applied on every element of a matrix, e.g., matrix addition, elementwise product
- other important operators and functions for matrices.
- matrix multiplication `%*%`, inverse of an invertible matrix `solve()`, matrix transposition `t()`

```{r matrix operators}

# contsruct the matrix
x <- matrix(c(2,1,3,4), nrow = 2, ncol = 2, byrow = TRUE)

# transpose a matrix
y <- t(x)

# matrix multiplication of x and y
x %*% y

# inverse of matrix x
z <- solve(x)

# matrix multiplication of x and inverse x
x %*% solve(x)
## if we get a weird rounding number here, use `round()`

```

We can extract elements from the matrix:

```{r extracting-elements}
# construct the matrix
new_mat1 <- matrix(1:16, nrow = 4, ncol = 4, byrow = TRUE)

# select the single element living at coordinates (2,3), or row 2, column 3
new_mat1[2,3]

# select a vector of elements, those living at (2,3), (3,3), and (4,3) or the entire 3rd column
new_mat1[2:4,3]

# select a submatrix of the original matrix
new_mat1[2:4, 2:3]

# if we don't supply an index to the second argument, it will return all indices
new_mat1[2:4,]

```

### Arrays

An array is a collection of elements stored in multiple dimensions. It contains elements with the same data type.

```{r array}
# construct an array with dimension 2x4x2, each element filled with 1
my_array <- array(1, dim=c(2,4,2))
my_array

# extract the (1,2,2)th element of the array, which is 1.
my_array[1,2,2]

```

### Data Frames

Most commonly used data structure to store data in R.

- similar to a matrix, but only requires each column to be elements of the same data type (each column is a valid vector).
- typically, each row of a df is an observation, and each column is a variable

```{r data-frames}

# constructing a data frame using `data.frame()`
## create the variable vectors with associated observations
p.height <- c(180, 155, 160, 167, 181)
p.weight <- c(65, 50, 52, 58, 70)
p.names <- c("Joanna", "Charlotte", "Helen", "Karen", "Amy")

## using the `data.frame()`, we're passing the name of the variable to the data frame
dataf <- data.frame(height = p.height, weight = p.weight, names = p.names)

# `colnames()` will tell us the variables (names) of the columns
colnames(dataf)
# `rownames()` will tell us the number of observations/records in the dataframe
rownames(dataf)

# we can change the names of the columns/rows
rownames(dataf) <- p.names

```

#### Extracting from dataframes

We can extract elements from a data frame similar to matrices, we use the syntax:

`[dataName]$[columnName]` to extract a column vector of a data frame, where `$` indicates accessing a property of a data frame object

We can also order a data frame similar to matrices

We can also add columns or rows using `rbind()` and `cbind()`

```{r extraction-df}
# create dataframes 1 and 2
df1 <- data.frame(id = 1:4, height = c(120, 150, 132, 122), weight =
c(44, 56, 49, 45))
df2 <- data.frame(id = 5:6, height = c(119, 110), weight = c(39, 35))

# append additional rows to a new dataframe
df_rcomb <- rbind(df1, df2)

# append additional columns to a new dataframe
df3 <- data.frame(id = 1:4, height = c(120, 150, 132, 122), weight =
c(44, 56, 49, 45))
df4 <- data.frame(location = c("UK", "CZ", "CZ", "UK"))
df_ccomb <- cbind(df3, df4)

# starting from an empty dataframe
df_example <- data.frame()
df_example <- rbind(df_example, df1)

```

### Factors

Factors are a special data type in R, like numeric, char, logical

- very useful to encode the group levels of observations. The most common way to construct a factor vector is not by concatenating factor data points, but by:

1. converting a char (or numeric) vector as a factor vector using the `as.factor()` function,
2. constructing a factor vector directly from a char (or numeric) vector using the `factor()` function

```{r factors}
# convert a numeric vector (below) as a factor vector with the levels vector `c("1","2","3")`, using `as.factor()`
# the levels vector is the character version of the unique set of values of the original vector, sorted into increasing order of it.
as.factor(c(3,1,2,1,2,3,2,1))

# using `factor()`
factor(c(3,1,2,1,2,3,2,1))

# or, we can use the `as.factor()` function with character values
as.factor(c("small", "medium", "large"))

# we can sort by the factor using `sort()` function
fact_x <- as.factor(c(3,1,2,1,2,3,2,1))
sort(fact_x)

# `order()` function will tell you the index, or location, of the elements, based on their factor, in the vector
order(fact_x)
```

If you want to change the order of levels of a factor, just use the `factor()` function and specify the `levels` argument

```{r factor-levels}

# construct a new factor vector from a numeric vector, and manually specify the levels of the factor as "2", "1", "3"
factor(c(1,1,1,2,2,3,3,3), levels = c(2,1,3))

# if we have too many levels than there are matching observations, it won't do anything
factor(c(1,1,1,2,2,3,3,3), levels = c(2,1,3,4))

# if we don't specify enough levels for the number of observations, it will provide NAs for the observations not specified
factor(c(1,1,1,2,2,3,3,3), levels = c(2,1))

```

### Graphics and plotting in R

#### Scatterplots

- plot the values of a numeric vector y against the index using `plot(y)`
- plot the values of a numeric vector y against another numeric vector x: `plot(x=x, y=y)`

Useful arguments:

- `xlab` = a title for the x-axis
- `ylab` = a title for the y-axis
- `main` = an overall title for the plot
- `type` options:
    - "p" for points
    - "l" for lines
    - "b" for both
    - "o" for both 'overplotted'
- `col` = color of the points and lines
- `lty` = line type, 1 for solid, 2 for dashed, 3 for dotted
- `xlim` = manually set the limits of the x-axis
- `ylim` = manually set the limits of the x-axis

```{r plotting-example}
# Import the base dataset “speed and stopping distances of cars”. Run ?cars for the description.
(data(cars))

# Scatterplot of the speed vector.
plot(cars$speed, ylab = "speed (mph)")

# Scatterplot of the distance vector against the speed vector.
plot(x=cars$speed, y=cars$dist, xlab="speed (mph)", ylab="stopping distance (ft)", main="cars data")

# Scatterplot of the distance vector against the speed vector, with lines connecting the data points.
plot(x=cars$speed, y=cars$dist, type="o", xlab="speed (mph)", ylab="stopping distance (ft)", main="cars data")

# If we fit a linear model, the fitted line is y = −17.579 + 3.932x. Add the straight lines `y = −17.579 + 3.932x` to the plot.
lm(cars$dist~speed$dist)
abline(a=-17.579, b=3.932, col = "red", lty = 2)

```

#### Histograms

- Plot the frequency histogram of a numeric vector x: `hist(x)`
    - we can manually specify the breakpoints in the histogram using the `breaks` argument
- plot the probability density histogram of a numeric vector x: `hist(x, freq=FALSE)`

```{r histograms}

# frequency histogram of the distance vector
hist(cars$dist)

# frequency histogram of the distance vector, with breakpoints manually specified as 0,10...120
hist(cars$dist, breaks = seq(from = 0, to = 120, by = 10))

# probability density histogram of the distance vector
hist(cars$dist, freq=FALSE)

# add a kernal density curve on the probability density histogram.
# we can also use the `lines()` function to add various kinds of lines to a plot

```

#### Boxplots

We can use the boxplot function create boxplots. From boxplots, we can:

1. graphically summarise the distribution of a variable
2. Identify potential unusual values,
3. Compare distributions between different groups

```{r boxplots}

# create a boxplot of the distance vector
boxplot(cars$dist)

# first, determine the mean speed `mean(cars$speed)`, determine where cars$speed is greater than the mean speed `cars$speed > mean(cars$speed)`, convert that result from a logical TRUE/FALSE to a numeric vector `as.numeric(cars$speed > mean(cars$speed))`, then group based on that and append it to the cars2 df:
cars2 <- cbind(cars, groups = as.numeric(cars$speed > mean(cars$speed)))

# create a boxplot of the new cars2 df, distance grouped by new group vector (speeding greater than average or not)

```

#### Pairs Plots

- we can use the `pairs()` function to create a multi-panel scatterplot (sometimes called a scatterplot matrix) which plots all combinations of variables


