---
title: "Multivariate Regression Lab 4"
author: "Matthew Reyes"
date: "`r Sys.Date()`"
output: html_document
---

# Multivariable Regression Lab 4

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 4 Notes

Graphics with ggplot (5.1, 5.2)
More on data frames (3.5)

### ggplot Graphics

1. Using the `flower.xls` dataset

```{r importing}

# libraries/packages
libraries <- c(
  "tidyr",
  "readr",
  "dplyr",
  "knitr",
  "ggplot2",
  "glmnet",
  "readxl"
)
lapply(libraries, require, character.only = TRUE)

# import data
flowers <- read.excel("data/flower.xls")

```

2. generating the plot

```{r generate-plot}

# includes the data and mapping layers
ggplot(mapping = aes(x = weight, y = shootarea), data = flowers)

# include a geom layer and lines
ggplot(aes(x = weight, y = shootarea), data = flowers) +
    geom_point() +
    geom_line()

# include a smoothed line
# geom_smooth() with color in the aes() of the geom_smooth() function
ggplot(aes(x = weight, y = shootarea), data = flowers) +
    geom_point() +
    geom_smooth(aes(colour = nitrogen), method = "lm", se = FALSE)

# moving color to the ggplot function now colors the data points associated with the fitted smooth lines
# black line is now a fitted lm for all the data points
ggplot(aes(color = nitrogen, x = weight, y = shootarea), data = flowers) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(method = "lm", se = FALSE, color = "black")

```

3. Customizing our plot further

```{r custom-plot}

# Further customizing our plot
ggplot(aes(color = nitrogen, x = weight, y = shootarea), data = flowers) +
    geom_point(aes(shape = nitrogen), size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(method = "lm", se = FALSE, linetype = 2, alpha = 0.6, color = "black") +
    geom_hline(aes(yintercept = mean(shootarea)), size = 0.5, colour = "black", linetype = 3) +
    xlab("Weight of flower (g)") + # X-axis label
    ylab("Area of shoot"~(cm^2)) + # Y-label axis
    labs(
        shape = "Nitrogen\nConcentration",
        color = "Nitrogen\nConcentration",
        caption = "Regression assumptions are unvalidated") +
    scale_colour_manual(values = c("#5c1aae", "#ae5c1a", "#1aae5c"),
        labels = c("High","Medium","Low")) +
    scale_shape_manual(values = c(15, 17, 19),
        labels = c("High","Medium","Low"))

# Further customizing our plot, now with factored nitrogen to specify the order of levels by concentration
flowers$nitrogen <- factor(flowers$nitrogen, levels = c("high", "medium", "low"))

# plot
ggplot(aes(color = nitrogen, x = weight, y = shootarea), data = flowers) +
    geom_point(aes(shape = nitrogen), size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(method = "lm", se = FALSE, linetype = 2, alpha = 0.6, color = "black") +
    geom_hline(aes(yintercept = mean(shootarea)), size = 0.5, colour = "black", linetype = 3) +
    xlab("Weight of flower (g)") + # X-axis label
    ylab("Area of shoot"~(cm^2)) + # Y-label axis
    labs(
        shape = "Nitrogen\nConcentration",
        color = "Nitrogen\nConcentration",
        caption = "Regression assumptions are unvalidated") +
    scale_colour_manual(values = c("#5c1aae", "#ae5c1a", "#1aae5c"),
        labels = c("High","Medium","Low")) +
    scale_shape_manual(values = c(15, 17, 19),
        labels = c("High","Medium","Low"))

# plot with 4 quadrants by treatment group
ggplot(aes(color = nitrogen, x = weight, y = shootarea), data = flowers) +
    geom_point(aes(shape = nitrogen), size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(method = "lm", se = FALSE, linetype = 2, alpha = 0.6, color = "black") +
    geom_hline(aes(yintercept = mean(shootarea)), size = 0.5, colour = "black", linetype = 3) +
    xlab("Weight of flower (g)") + # X-axis label
    ylab("Area of shoot"~(cm^2)) + # Y-label axis
    labs(
        shape = "Nitrogen\nConcentration",
        color = "Nitrogen\nConcentration",
        caption = "Regression assumptions are unvalidated") +
    scale_colour_manual(values = c("#5c1aae", "#ae5c1a", "#1aae5c"),
        labels = c("High","Medium","Low")) +
    scale_shape_manual(values = c(15, 17, 19),
        labels = c("High","Medium","Low")) +
    facet_grid(~ block + treat)

# plot with 4 quadrants by treatment group, own graphs (basically a 2x2)
# displays tip type crossed with treatment groups, 1 or 2
ggplot(aes(color = nitrogen, x = weight, y = shootarea), data = flowers) +
    geom_point(aes(shape = nitrogen), size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(method = "lm", se = FALSE, linetype = 2, alpha = 0.6, color = "black") +
    geom_hline(aes(yintercept = mean(shootarea)), size = 0.5, colour = "black", linetype = 3) +
    xlab("Weight of flower (g)") + # X-axis label
    ylab("Area of shoot"~(cm^2)) + # Y-label axis
    labs(
        shape = "Nitrogen\nConcentration",
        color = "Nitrogen\nConcentration",
        caption = "Regression assumptions are unvalidated") +
    scale_colour_manual(values = c("#5c1aae", "#ae5c1a", "#1aae5c"),
        labels = c("High","Medium","Low")) +
    scale_shape_manual(values = c(15, 17, 19),
        labels = c("High","Medium","Low")) +
    facet_grid(block ~ treat) +
    theme_classic()

# Adding a theme function to our graph
theme_rbook <- function(
    base_size = 13, 
    base_family = "", 
    base_line_size = base_size/22, 
    base_rect_size = base_size/22 ) {
        theme(
            axis.title = element_text(size = 13),
            axis.text.x = element_text(size = 10),
            axis.text.y = element_text(size = 10),
            plot.caption = element_text(size = 10, face = "italic"),
            panel.background = element_rect(fill="white"),
            axis.line = element_line(size = 1, colour = "black"),
            strip.background = element_rect(fill="#cddcdd"),
            panel.border = element_rect(colour = "black", fill=NA),
            strip.text = element_text(colour = "black"),
            legend.key  = element_blank()
        )
}

# apply our new theme to the graph
ggplot(aes(color = nitrogen, x = weight, y = shootarea), data = flowers) +
    geom_point(aes(shape = nitrogen), size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_smooth(method = "lm", se = FALSE, linetype = 2, alpha = 0.6, color = "black") +
    geom_hline(aes(yintercept = mean(shootarea)), size = 0.5, colour = "black", linetype = 3) +
    xlab("Weight of flower (g)") + # X-axis label
    ylab("Area of shoot"~(cm^2)) + # Y-label axis
    labs(
        shape = "Nitrogen\nConcentration",
        color = "Nitrogen\nConcentration",
        caption = "Regression assumptions are unvalidated") +
    scale_colour_manual(values = c("#5c1aae", "#ae5c1a", "#1aae5c"),
        labels = c("High","Medium","Low")) +
    scale_shape_manual(values = c(15, 17, 19),
        labels = c("High","Medium","Low")) +
    facet_grid(block ~ treat) +
    theme_rbook()
```

### Data Frames

- `dim`/`nrow`/`ncol` functions return the dimension, number of rows, number of cols of a data frame (or an array
- `head`/`tail` functions:
    - return the first/last several rows of a data frame (or first/last parts of an object)
- `view` function: invoke a data viewer
- `summary`: summarize a data frame or an object
- `str`: display the structure of a data frame or object
- `table`: build a contingency table of the counts at each combination of factor levels

```{r tables}

# return a vector of counts at each treat level
table(flowers$treat)

# return a matrix of counts at each combination of the treat and nitrogen levels
table(flowers$treat, flowers$nitrogen)

# return an array of counts at each combination of the treat and nitrogen levels
table(flowers$treat, flowers$nitrogen, flowers$block)

# it's fine if one variable is continuous
table(flowers$treat, flowers$flowers)


```

apply function: apply functions over each row or each column of a data frame (or over
array margins).

```{r apply-function}

apply(flowers[,4:8], 2, mean)
#Apply the mean function over each column of flower[,4:8].
apply(flowers[,4:8], 1, mean)
#Apply the mean function over each row of flower[,4:8].
data("HairEyeColor")
apply(HairEyeColor, 1, mean)
#Apply the mean function over the first dimension of HairEyeColor.
apply(HairEyeColor, 2, mean)
#Apply the mean function over the second dimension of HairEyeColor.
apply(HairEyeColor, c(1,2), mean)
#Apply the mean function over the first and second dimension of HairEyeColor.

```

Check the usage of the lapply, sapply, tapply functions if you are interested.

### Distributions

Normal and binomial distributions as examples, along with generating random numbers

```{r randoms}

# generate 100 random numbers from the normal distribution N(1,1.2^2) with mean 1 and variance 1.2
x <- rnorm(100, mean = 1, sd = 1.2)

# check our stats of x
mean(x)
var(x)

# generate 100 random numbers from the binomial distribution B(10, 0.7). Remember, the binomial distribution B(n,p) has mean np and variance np(1-p)
x_binom <- rbinom(100, size = 10, prob = 0.7)

# check our binom stats
mean(x_binom)
var(x_binom)

```

Computing a cumulative distribution function. For random variable X, the cumulative dist. function F(x) of the probability distribution is the probability of X < x.

```{r cumulative-dist}

# returns 0.5, because P(X < 0) = 0.5 if X ~ N(0, 1)
pnorm(0, mean = 0, sd = 1)

# returns 0.9750021, because (see pdf)
pnorm(1.96, mean = 0, sd = 1)

# check by running integrate fxn
integrate(dnorm, lower = -100, upper = 1.96)

# returns 0.04734899, because (see pdf)
pbinom(4, size = 10, prob = 0.7)

# can verify by running:
sum(choose(10, c(0:4))*0.7^c(0:4)*0.3^c(10:6))

```

Probability density function or the probability mass fxn. For one continuous random var X, the probability density function f(x) satisfies P( X < x ) = integral( f(x)dx ). 

For discrete random var X, the prob mass function m(x) is the probability of X = x.

```{r probability-functions}

dnorm(0, mean = 0, sd = 1)

# verify:
1/sqrt(2*pi)*exp(0)


dbinom(4, size = 10, prob = 0.7)

# verify:
choose(10,4)*0.7^4*0.3^6

```

Quantile function. Informally, the quantile function is the inverse of the cumulative distribution function

```{r quantile-function}

# returns 1.959964. This is because P( X , 1.959964 ) ~~ 0.975 if X ~ N (0,1)
qnorm(0.975, mean = 0, sd = 1)

# binom function, returns 5
qbinom(0.04734899, size = 10, prob = 0.7)

# returns 4, because 0.04734898 < P( X < 4 ) < 0.04734899 < P( X < 5 ), if X ~ B(10, 0.7)
qbinom(0.04734898, size = 10, prob = 0.7)

```